A Review of Mortality Models

### 2020-09-08 mortality-models/draft.tex
WIP 08Sep20

\documentclass{article}
\usepackage{amsmath, amssymb, amsthm, amsfonts}
\numberwithin{equation}{section}
\DeclareMathOperator{\pr}{P}

\title{A Review of Mortality Models}
\author{To be filled}
\date{Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s0}
Although it is very difficult to predict the behaviour of a member of a 
population one can often make statistical statements with much greater 
confidence. For example, it is impossible to predict when a particular 
radioactive atom will decay but one can make an accurate prediction of how 
many atoms will decay in a long enough time interval. Likewise, it is not
possible to predict the time of death of an individual although one reliably
predict how many people from a large enough group will survive past a certain
date in future. Actuaries knew this statistical fact long before the 
mathematical foundations of the theory of probability were laid. The 
mathematical models they to describe how deaths occur in a given society are
called `mortality models'.

The death of an individual in a population or the decay of a radioactive atom
is an event happening to a member of the population. Likewise, the failure of
a machine, an occurrence of an event, the duration of a certain phase in an
economy are all different manifestations of the same phenomenon in which 
something happens that significantly alters the future of the member of the
population. Death terminates an individual's existence, radioactive decay
changes the atom, failure renders a machine unusable and economies change 
after occurrences of certain cataclysmic events. Mortality of an individual
is akin to reliability of a machine. Life of an individual is similar to
the duration of a certain phase of business cycle. As a result, the 
theoretical tools used to study these phenomena are similar to each other. 
Mortality analysis in actuarial science is called survival analysis is 
statistics, reliability theory in engineering, duration analysis in economics 
and event history analysis in sociology.

In this article we shall review the development mathematical models to study
mortality in large enough populations of people.

@@

\section{Other mortality laws prior to the 20th century}\label{s4}
de Moivre's, Gompertz's and Makeham's laws of mortality were not the only
ones in vogue in the nineteenth century and prior to it. In this section we
mention a few notable ones without commenting on their merits. Lambert's
survival function \cite{lambert1776dottrina} was
\begin{equation}\label{s4e1}
s(x) = \left(\frac{a - x}{x}\right)^2 - b\left(e^{-x/c} - e^{-x/d}\right).
\end{equation}
It had four degrees of freedom. Babbage\cite{babbage1823tables} proposed a 
quadratic survival function
\begin{equation}\label{s4e2}
s(x) = c - bx - ax^2.
\end{equation}
Young proposed a polynomial of degree $40$ as the survival function in 
\cite{adler1866memoir} 1826. von Littrow\cite{von1832lebensversicherungen}
extended Babbage's model to polynomials of higher degree. Moser
\cite{moser1839gesetze} proposed a polynomial survival function with 
fractional powers of $x$. It has five degrees of freedom and had the form
\begin{equation}\label{s4e3}
s(x) = 1 - ax^{1/4} _ bx^{9/4} - cx^{17/4} - dx^{25/4} + ex^{33/4}.
\end{equation}
All these models were attempts to fit the mortality data with a continuous 
function. They were not rooted in demographic observations like the models
of Gompertz, Makeham, Lazarus or Thiele. As a result, they soon fell into
disuse after Gompertz proposed his model and others improved upon it.

@@

\bibliography{draft}
\bibliographystyle{plain}
\end{document}

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, amsthm, amsfonts}
\numberwithin{equation}{section}
\DeclareMathOperator{\pr}{P}
\renewcommand{\vec}[1]{\mathbf{#1}}

\title{A Review of Mortality Models}
\author{To be filled}
\date{Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s0}
Although it is very difficult to predict the behaviour of a member of a large
population one can often make statistical statements about the same population 
with much greater 
confidence. For example, it is impossible to predict when a particular 
radioactive atom will decay but one can make an accurate prediction of how 
many atoms will decay in a long enough time interval. Likewise, it is not
possible to predict the time of death of an individual although one can reliably
predict how many people from a large enough group will survive past a certain
date in future. Actuaries knew this statistical fact long before the 
mathematical foundations of the theory of probability were laid. The 
mathematical models they to describe how deaths occur in a given society are
called `mortality models'.

The death of an individual in a population or the decay of a radioactive atom
is an event happening to a member of the population. Likewise, the failure of
a machine, an occurrence of an event, the duration of a certain phase in an
economy are all different manifestations of the same phenomenon in which 
something happens that significantly alters the future of the member of the
population. Death terminates an individual's existence, radioactive decay
transforms the atom, failure renders a machine unusable and economies change 
after occurrences of certain cataclysmic events. Mortality of an individual
is akin to reliability of a machine. Life of an individual is similar to
the duration of a certain phase of business cycle. As a result, the 
theoretical tools used to study these outwardly different phenomena are similar 
to each other. 
Mortality analysis in actuarial science is called survival analysis is 
statistics, reliability theory in engineering, duration analysis in economics 
and event history analysis in sociology.

John Graunt, an English haberdasher, was perhaps the world's first demographer
and epidemiologist. In spite of the rudimentary mathematical tools available
in the seventeenth century, he was able to publish the first life tables. They
were based on the `Bills of Mortality' published in the parishes of London
every Thursday. The data was not best suited for constructing life tables but
that was the only one available in Graunt's time. Graunt admitted the 
inadequacy of his approach. Edmund Halley, the man after whom a comet is 
named, could get demographic data of a much better quality from the city of
Wroclaw in Poland. It had the age and sex of the dead individuals. Halley
used this data to build life tables of a greater accuracy than Graunt's. He
also demonstrated how to use the life tables to compute annuities. Graunt's
and Halley's work was entirely empirical, keeping with the times in which
it was done. Using their life tables was also quite laborious. With the 
invention of Calculus in the late seventeenth century and the development of 
the rudimentary ideas of a mathematical function, it became possible to build
theoretical models of human mortality.

Traditionally, mortality models studied existing demographic data to produce
mortality tables. They described mortality at a fixed point in time. The pace 
of technology development was quite slow in the times when the first mortality
models were proposed. Therefore, the mortality data and the models describing
them appeared to be like a natural characteristic of the human species. They
sufficed the needs of the fledgling insurance industry to compute projections
of annuities.

However, the advancement in health science since the early years of the 
twentieth century, the relative peace in the western world since the second 
world war and abundance of food in the rich world resulted in a change in
historical mortality patterns. Pension funds started realizing that annuities
were being drawn for longer than expected periods. As the population in the
developed countries started greying the pension funds had the additional 
problem of a higher annunity obligations and a lower contribution from newer
workers. Pension funds now have a keen interest in understanding how mortality 
rates will evolve in the future. They now want to forecast mortality than just
produce models of the current mortality.

Demographic forecasting predates mortality forecasting. Economists like Thomas
Malthus predicted that the human population will grow to such an extent that
there will not be enough to feed everyone. This dire prediction meant that the
politicians and the economists had a keen desire to predict how populations
will grow and whether there will be enough food for everyone. As a result, many 
statistical methods developed in Demography were readily available for mortality
forecasting. This need for forecasting was also coeval with the developments
in the modern theory of mathematical statistics and the foundations of
probability. It is not a coincidence that the early mortality models were
mathematical functions while the later ones are stochastic models.

Mortality models, being mathematical functions, produce a single curve for
a given set of its parameters. On the other hand, mortality forecasts are a
collection of curves or a family of probablity distributions. Their output
is accompanied with appropriate confidence intervals to indicate the 
uncertainty in their reported values.

In this article we shall review the development mathematical models to study
and forecast mortality in large enough populations of people.

@@

\section{Other mortality laws prior to the 20th century}\label{s4}
de Moivre's, Gompertz's and Makeham's laws of mortality were not the only
ones in vogue in the nineteenth century and prior to it. In this section we
mention a few notable ones without commenting on their merits. Lambert's
survival function \cite{lambert1776dottrina} was
\begin{equation}\label{s4e1}
s(x) = \left(\frac{a - x}{x}\right)^2 - b\left(e^{-x/c} - e^{-x/d}\right).
\end{equation}
It had four degrees of freedom. Babbage\cite{babbage1823tables} proposed a 
quadratic survival function
\begin{equation}\label{s4e2}
s(x) = c - bx - ax^2.
\end{equation}
Young proposed a polynomial of degree $40$ as the survival function in 
\cite{adler1866memoir} 1826. von Littrow\cite{von1832lebensversicherungen}
extended Babbage's model to polynomials of higher degree. Moser
\cite{moser1839gesetze} proposed a polynomial survival function with 
fractional powers of $x$. It has five degrees of freedom and had the form
\begin{equation}\label{s4e3}
s(x) = 1 - ax^{1/4} + bx^{9/4} - cx^{17/4} - dx^{25/4} + ex^{33/4}.
\end{equation}
All these models were attempts to fit the mortality data with a continuous 
function. They were not rooted in demographic observations like the models
of Gompertz, Makeham, Lazarus or Thiele. As a result, they soon fell into
disuse after Gompertz proposed his model and others improved upon it.

@@

\section{Lee-Carter model}\label{s6}
The problem of forecasting population changes and composition is of interest
to both the demographers and the pension fund managers. Ronald Lee and 
Lawrence Carter's motivation for developing a model to forecast mortality was
the observation that the human life expectancy rose from $47$ years to $75$ 
years in the United States in the $88$ years starting from the year $1900$.
Should this trend continue unabated in the future, the human life expectancy
will touch $100$ years by the year $2065$ upsetting the Social Security 
Administration's calculations based on the long-term life expectancy of about
$80.5$ years. The wide difference between Lee and Carter's 
\cite{lee1992modeling} forecasts and the assumptions of the pension funds 
made the actuaries turn their attention to forecasting than modelling alone.

Unlike the previous models which tried to build formulae that mimicked the
behaviour of human mortality in various stages of life, Lee and Carter used
the statistical techniques of time series for their projection into the future.
They excluded factors representing advancement in medicine or the behavioural
and social changes on mortality. Their time series model extrapolated the 
long-term trends seen in demographic data over several decades. They did not
find it necessary to assume that the human life cannot extend beyond a certain
limit. Nothing analogous to the $\omega$ in de Moivre's model would be found
in Lee and Carter's.

The mathematical form of Lee and Carter's model is
\begin{equation}\label{s6e1}
\log{M}(x, t) = {a}_x + {b}_x{k}_t + \epsilon_{x, t},
\end{equation}
where the coefficients ${a}_x$ and ${b}_x$ depend on the age, the
`index' ${k}_t$ depends on time and ${M}(x, t)$ is the mean death
rate for age $x$ in time $t$. Since
\[
\frac{d}{dt}{M}(x, t) = {b}_x\frac{dk_t}{dt},
\]
${b}_x$ captures the the changes in mortality rates due to changes in
the index with time. A positive value of a component of ${b}_x$ indicates
falling mortality while a negative value indicates the opposite. The term
$\epsilon_{x, t}$ is an `error term' with mean $0$ and variance that depends
on the age. It captures the factors that influence mortality but are not
explicitly included in the model.

The model of equation \eqref{s6e1} does not have a unique solution. If 
${a}$, ${b}$ and ${k}$ form one solution then for any number $c$,
the triple ${a} - c{b}, {b}, {k} + c$ is also a solution. In
fact, even ${a}, c{b}, {k}/c$ is a solution if $c \ne 0$. This phenomenon is
also called `identifiability problem'\cite{plat2009stochastic}.

Fitting the Lee-Carter model consists in finding the parameters ${a}_x,
{b}_x$ and ${k}_t$ for a given matrix of mortality rates ${M}(x, t)$.
The model is solved using the singular value decomposition (SVD). The algorithm
is described in appendix A to Lee and Carter's paper\cite{lee1992modeling} 
introducing the model.

Lee and Carter's forecast of life expectancy for the American population was 
$86.05$ by the year $2065$. The Actuary of the Social Security Administration,
on the other hand, forecast it as $80.5$. In $2020$ the observed value is 
$78.87$, suggesting that Lee and Carter's forecast might be closer to the 
reality in the year $2065$.

The advantages of Lee-Carter model lie in its robustness, simplicity and 
the fact that it fits well over wide ranges of age. It is also observed to be
able to fit small non-linearities in the mortality curve quite well. Its 
shortcomings are its lack of smoothness of age effects in small populations, 
inability to work with cohorts, inability to cope with improvements varying 
across times and a possible underestimation of uncertainty. It also has a 
limited versatality owing to it being a one-factor model.

\section{Age-Period-Cohort models}\label{s7}
Demographic data is often analysed by taking into account the age, period 
and the cohort of individuals. The analysis need not be for mortality alone.
It could be for incidence of diseases or even economic consumption. Consumption
patterns do depend on age of the individual. But when they are observed over
a long enough duration they are also found to depend on the period and the 
cohort. For instance, certain goods may not have been available or would have
been expensive in a certain period and therefore we not consumed by many. It
is also well-known that different age groups of people have different tastes.
The age-period-cohort (APC) models are a class of additive models where the
predictor is a sum of three time effects, age, period and cohort. APC models
are thus able to take into account the impact of the changes in the society 
and the available technology.

Period analysis alone is quite useful to understanding the effect of wars,
pandemics, revolutions or other upheavals on a population. However, they
capture the effect of the times in which people are living. Behavioural
changes in people are better captured by considering cohorts. The 
characteristics of a cohort are an aggregate of the characteristics of its 
members\cite{willekens1991age}. Modern APC models require longitudnal data
to understand the life processes as they evolve. They do not consider just
the event of the death of an individual but the process of dying whose 
culmination is death. They take the view that the mortality risks in a 
population are better determined by an understanding of morbidity in it
\cite{van1990determinants}.

The APC framework is used widely in the social sciences. In the context of
mortality modelling and forecasting it started being used while answering the
questions \cite{willekens1991age}:
\begin{itemize}
\item Does mortality show a greater regularity when viewed at the level of a
cohort or for a particular period? The answer to this question is important
for mortality forecasting.

Although a full-fledged APC framework arose relatively recently, the importance
of cohort analysis was not missed by early demographers and actuaries. Derrick
\cite{derrick1927observations} observed in 1927 that the ratio of mortality
for one cohort to that of another was constant for all ages above $10$ years.
It suggested that it may be possible to forecast the mortality of an individual
based on the cohort in which he falls. It was also observed that the mortality
rates for the generation that was young during the years of the second world
war were overestimated for obvious reasons. This once again indicates the 
importance and the role of cohorts in modelling mortality.

\item Do events early in life affect the experiences later in life? This 
question can be answered only in the context of a cohort analysis.

While studing the death rates in England, Scotland and Sweden, Kermack
\cite{kermack1934death} observed that the differences in mortality was mostly
because of the environmental conditions experienced in the first fifteen years
of life. It was not a result of the other influences in later life. The 
study also uncovered the fact that an improvement in the conditions around
child-birth boosted survival rates of new-borns. It proposed that the health
of the mother has a positive impact on the health of the child and its chances
to live longer. Preston and de Walle \cite{preston1978urban} came to similar
conclusion for the French mortality data. This idea was further supported by 
Caselli and Capocaccia's study\cite{caselli1989age} on Italian demographics 
that infant mortality had a weakening effect on eventual mortality. However, 
this observation is not universally accepted. Other studies indicate that a 
high infantile mortality results in lower eventual mortality because the 
`stronger' individuals stay long enough to live long enough 
\cite{manton1981methods}. Lest one concludes that only childhood environment
determines mortality rate later in life, Horiuchi reported 
\cite{horiuchi1983long} that the malnutrition and physical privation during the
second world war wreaked the health of those who lived through it and affected
their mortality when they became old.
\end{itemize}

APC studies are plagued by what is called the `identification problem' common
to all cohort analysis. It is the problem of assigning an individual to the
right cohort when the a) the cohorts are adjacent in time, b) the individual 
is on the border line of two adjacent cohorts and c) the characteristics of
the cohorts are quite distinct. The problem can be mitigated by following
the directions suggested in Willekens and Scherbov's \cite{willekens1991age}
analysis.

APC model can be expressed as a generalized linear model in which we predict
the mortality rate from age, period and cohort as predictors. We usually use
the variables in the data that are the most appropriate proxy for age, period
and cohort. Once the model is build from the available data, we can use it to 
forecast mortality for a newer set of predictors. It is assumed that the 
mortality rate can be described as a Poisson process. This assumption is found 
to be valid for societies with low death rates. It may not hold good when 
the society is experiencing a war or a pandemic.

\section{Cairns-Blake-Dowd model}\label{s8}
Cairns, Blake and Dowd proposed a model \cite{cairns2006two} to study the
evolution of mortality beyond sixty years of age and its impact on longevity
risk. It was a two factor stochastic model. Of these, the first factor affects
the mortality rate at all ages in an identical manner and the second one 
affects the advanced years to a greater extent. Their mortality curve is
defined in terms of forward survival probability $\pr(t, T_0, T_1, x)$. It is
the probability measured at $t$ that an individual aged $x$ at time $0$,
alive at $T_0$ will survive until time $T_1 > T_0$. The curve is represented
by the equation
\begin{equation}\label{s8e1}
1 - p(t+1,t,t+1,x) = 
\frac{e^{A_1(t+1)+A_2(t+1)(x+t)}}{1+e^{A_1(t+1)+A_2(t+1)(x+t)}},
\end{equation}
where $A_1(u)$ and $A_2(u)$ are stochastic processes. If $q(p, T_0, T_1, x)
= 1 - p(t, T_0, T_1, x)$ then we can readily write equation \eqref{s8e1} in
the form
\begin{equation}\label{s8e2}
\log\left(\frac{q(t,T_0,T_1,x)}{p(t,T_0,T_1,x)}\right) = 
A_1(t+1) + A_2(t+1)(x+t).
\end{equation}
The log-odds of probability of death is thus a linear function of age and the
coefficients of this relation vary with time. The model uses data from 
different cohorts at a common point in time. Therefore, the coefficients 
$A_1(t)$ and $A_2(t)$ are interpreted as factors. The original model of Cairns,
Blake and Dowd used ordinary least square regression to estimate the factors.
The authors plotted the trend of the factors for the period between $1961$ and
$2002$. The factor $A_1(t)$ decreases with time indicating an improvement in
the mortality. On the other hand, $A_2(t)$ is observed to increase with time.
This suggests that the improvements in mortality are felt to a lesser extent 
for older people. In order to forecast the factors at times in future, the 
authors model $A(t) = (A_1(t), A_2(t))^T$ as a random walk with a drift. That
is,
\begin{equation}\label{s8e3}
A(t+1) - A(t) = \mu + CZ(t+1),
\end{equation}
where $\mu$ is a constant vector representing the mean increment, $C$ is a 
constant $2 \times 2$ upper triangular matrix and $Z(t)$ is a standardised
two dimensional gaussian random variable. The authors also propose a criterion
for a good model of this kind. They introduce the idea of `biological 
reasonability'. In general, a biologically reasonable model will have the 
first component of $\mu$ negative and the second component positive. Further,
the mortality rates for older cohorts should normally be greater than the 
younger cohorts.

Practitioners favour Cairns-Blake-Dowd model for its robustness and the ease
with which it allows inclusion of parameter uncertainty. It is also more 
versatile than the Lee-Carter model because it has an additional factor. 
However, like the Lee-Carter model it does not allow analysis of cohort effects.
Although it models a wide range of advanced years, it does not fit the 
mortality data as well as Lee-Carter model.

\section{Stochastic cohort models}\label{s9}
The stochastic models of Lee-Carter and Cairns-Blake-Dowd were unable to
incorporate cohort effects. To remedy this shortcoming, Renshaw and Haberman
\cite{renshaw2006cohort} proposed a modification of the Lee-Carter model that 
includes cohorts. It can be expressed mathematically as
\begin{equation}\label{s9e1}
\log{M}(x, t) = {a}_x + {b}_x{k}_t + {c}_x\gamma_{t-x},
\end{equation}
where the symbols have the same meaning as in Lee-Carter model. In the new
term, third on the right hand side, $\gamma_{t-x}$ captures the cohort effect.
It is a function of $t - x$, the (approximate) year of birth and ${c}_x$
is just the regression coefficient of this term. Although the model fitted the
male mortality data in England and Wales quite well it turned to not as 
robust as other models. The lack of robustness is believed to be related to
the shape of the likelihood function. It has more than one local maximum.
Therefore, when we change the age or the year range the optimizer ends up
settling on a different maximum each time\cite{cairns2008modelling}. Currie
simplified the Renshaw-Haberman model to
\begin{equation}\label{s9e2}
\log{M}(x, t) = {a}_x + {k}_t + \gamma_{t-x}.
\end{equation}
The simplification solved the problem of robustness but made the fit poorer
\cite{plat2009stochastic}. Plat introduced a model to address the deficiencies
of all the previous models while retaining their benefits 
\cite{plat2009stochastic}. Its mathematical structure is
\begin{equation}\label{s9e3}
\log M(x,t) = a_x + \kappa^1_t + \kappa^2_t(\bar{x} - x) + 
\kappa^3_t(\bar{x} - x)^+ + \gamma_{t-x},
\end{equation}
where $(\bar{x} - x)^+ = \max(\bar{x} - x, 0)$ and $\bar{x}$ is the mean age
in the sample range. The term $a_x$ plays a 
similar role as it did in Lee-Carter model. It controls the general shape
of the mortality curvve. Of the four stochastic factors, $\kappa_t^1$
represents the change in the level of mortality across all ages, $\kappa_t^2$
represents the change in the level of mortality between ages, $\kappa_t^3$
captures the differences in mortality behaviour in the early adulthood stage
and $\gamma_{t-x}$ captures the cohort effect. If one is interested in 
modelling only the advanced ages then one need not worry about the `accident
hump' of young adulthood and use the simplified model,
\begin{equation}\label{s9e4}
\log M(x,t) = a_x + \kappa^1_t + \kappa^2_t(\bar{x} - x) + \gamma_{t-x},
\end{equation}
Like the Lee-Carter model, this one too suffers from the `identifiability
problem'. For constants $c_1, c_2, d$, the set of parameters
\begin{eqnarray*}
\bar{\gamma}_{t-x} &=& \gamma_{t-x} + c_1 + c_2(t - x) \\
\bar{\kappa}_t^1 &=& \kappa^1_t + c_1 - d\bar{x}c_2 \\
\bar{\kappa}_t^2 &=& \kappa^2_t + dc_2 \\
\bar{a}_x &=& a_x + (1 - d)c_2
\end{eqnarray*}
also lead the the same values for $\log M(x, t)$. The problem is mitigated 
by requiring that 
\begin{eqnarray*}
\sum_{i=x_0}^{x_1} \gamma_i &=& 0 \\
\sum_{i=x_0}^{x_1} i\gamma_i &=& 0 \\
\sum_t\kappa_t^3 &=& 0,
\end{eqnarray*}
where $x_0$ and $x_1$ are the earliest and the latest years of birth to which
the cohort effect is fitted. 
\bibliography{draft}
\bibliographystyle{plain}
\end{document}
